Natural variation in teosinte at the domestication locus teosinte branched1 (tb1)

### 2014-04-15 manuscript.tex
edits
intro

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. Perhaps the central focus of the study of domestication has been the identification of genetic variation underlying agronomically important traits such as fruit size and plant architecture \citep{OlsenGross2010}. Additionally, many domesticates show reduced genetic diversity when compared to their wild progenitors, and an understanding of the distribution of diversity in the wild and its phenotypic effects has become increasingly useful to crop improvement \cite{KovachMcCouch2008}. But while some effort has been invested into understanding how wild alleles behave in their domesticated relatives \jri{CITE}, very little is known about the role that alleles found most commonly in domesticates play in natural populations of their wild progenitors.
\jri{CITE; perhaps Arias and Rieseberg 1994 TAG?}

Maize (\emph{Zea mays} ssp. \emph{mays}) was domesticated from the teosinte \emph{Zea mays} ssp. \emph{parviglumis} (hereafter, \emph{parviglumis}) roughly 10,000 B.P. in southwest Mexico \citep{Piperno2009, Matsuoka2002}. \jri{shouldn't this be 9K BP? what does Piperno 2009 say?} Domesticated maize and the teosintes are an attractive system in which to study domestication due to the abundance of genetic tools developed for maize and well-characterized domestication loci \citep{Hufford2012a, Doebley2004, Hufford2012b}. Additionally, large naturally occurring populations of both \emph{Zea mays} ssp. \emph{parviglumis} (the wild progenitor of maize) and \emph{Zea mays} ssp. \emph{mexicana} (highland teosinte; hereafter \emph{mexicana}) can be found throughout Mexico \citep{Wilkes1977, Hufford2013}, and genetic diversity and effective population size (Ne) of these taxa is estimated to be high \citep{Ross-Ibarra2009} relative to other plants \citep{Nybom2004}. \jri{what are nybom's estimates? i didn't think mine were that high and don't remember Nybom giving Ne estimates}

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis of these changes has been a focus of maize research for a number of years\citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \citep{Doebley1990, DoebleyStec1991, LukensDoebley1999}. \jri{an odd assortment of cites. why not the 1997 nature paper?} Further work showed that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \citep{Studer2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. Furthermore, \citep{StuderDoebley2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte. \jri{MBH todo}

>>>>

Domesticated crops and their wild progenitors provide an excellent system in which to study adaptation and genomic changes associated with human-mediated selection \citep{Ross-Ibarra2007}. Perhaps the central focus of the study of domestication has been the identification of genetic variation underlying agronomically important traits such as fruit size and plant architecture \citep{OlsenGross2010}. Additionally, many domesticates show reduced genetic diversity when compared to their wild progenitors, and an understanding of the distribution of diversity in the wild and its phenotypic effects has become increasingly useful to crop improvement \citep{KovachMcCouch2008}. But while some effort has been invested into understanding how wild alleles behave in their domesticated relatives \jri{CITE}, very little is known about the role that alleles found most commonly in domesticates play in natural populations of their wild progenitors.
\citep{Whitton1997}.

Maize (\emph{Zea mays} ssp. \emph{mays}) was domesticated from the teosinte \emph{Zea mays} ssp. \emph{parviglumis} (hereafter, \emph{parviglumis}) roughly 9,000 B.P. in southwest Mexico \citep{Piperno2009, Matsuoka2002}. Domesticated maize and the teosintes are an attractive system in which to study domestication due to the abundance of genetic tools developed for maize and well-characterized domestication loci \citep{Hufford2012a, Doebley2004, Hufford2012b}. Additionally, large naturally occurring populations of both \emph{Zea mays} ssp. \emph{parviglumis} (the wild progenitor of maize) and \emph{Zea mays} ssp. \emph{mexicana} (highland teosinte; hereafter \emph{mexicana}) can be found throughout Mexico \citep{Wilkes1977, Hufford2013}, and genetic diversity and effective population size (Ne) of these taxa is estimated to be high \citep{Ross-Ibarra2009} relative to other plants \citep{Nybom2004}. \jri{what are nybom's estimates? i didn't think mine were that high and don't remember Nybom giving Ne estimates}

Many morphological changes are associated with the domestication of maize, and understanding the genetic basis of these changes has been a focus of maize research for a number of years\citep{Doebley2004}. One of the most dramatic changes is found in plant architecture: domesticated maize is characterized by a central stalk with few tillers and lateral branches terminating in a female inflorescence, while teosinte is highly tillered and bears tassels (male inflorescences) at the end of its lateral branches. The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of organ growth, was identified as a major QTL involved in branching differences between maize and teosinte \citep{Doebley1990, DoebleyStec1991, LukensDoebley1999}. \jri{an odd assortment of cites. why not the 1997 nature paper?} Further work showed that the insertion of a 4.9 kb retrotransposon (Hopscotch) in the upstream control region of \emph{tb1} led to increased expression of this gene, causing the reduction in branching observed in domesticated maize \cite{Studer2011}. The effects of this insertion have been observed in tiller number in maize, but little is known about its role, if any, in teosinte \citep{Studer2011}. Dating of this element has suggested that its insertion predates the domestication of maize, leading to the hypothesis that it was segregating as standing variation in ancient populations of teosinte and increased to high frequency in maize due to selection during domestication \citep{Studer2011}. Furthermore, \citep{StuderDoebley2012} investigated the phenotypic effects of 9 teosinte \emph{tb1} alleles in an isogenic maize background and found that the introgressions sort into three distinct phenotypic classes, suggesting that variation at the \emph{tb1} locus may play a functional role in teosinte. \jri{MBH todo} \lev{hmm?}
